---
title: "A whirlwind tour of Quarto and the Tidyverse"
author: "Paul M. Magwene"
format: html
---

# Installing packaages

See in class demo on the Packages install dialog.

-   Install the tidyverse package

# Loading packages

Once a package is installed we can load it the the `library` function:

```{r}
library(tidyverse)
```

# Quarto

Quarto is a tool for scientific documentation and publishing. In this class we'll be using Quarto documents as the primary method for writing, testing, and exploring R code.

## In-class tour of Quarto

-   Create a Quarto document

-   Review Quarto interface in RStudio

-   Visual vs Source editor view

-   Code blocks

-   Interactive notebook model vs document rendering – discuss differences and possible points of confusion about execution models

    -   Take home: When testing your notebooks (e.g. homeworks!) use the `Session > Restart R and Run All Chunks` menu item to make sure your code runs without error

# Data import using the readr package

The `readr` package is one of the libraries included when we installed the tidyverse. The readr package provide a wide variety of functions for loading common data formats like CSV, TSV, and other delimited data types.



## Load a data file using `readcsv`

Download `penguins.csv`

```{r}
penguins <- read_csv("~/Bio724/penguins.csv")
```


# Simple data frame manipulations

When the penguins data set is loaded it is treated as a "data frame", which is an R data structure that represents tabular data. The data frame is the data structure we'll be using most often in this class.

Basic functions for getting a handle on what's in a data frame

-   `View`
-   `dim`
-   `head` and `tail`
-   `str`

# Basic plots

## Raw data plot and jittering

```{r}
ggplot(penguins, aes(x = body_mass_g, y = 0)) + 
  geom_point(alpha=0.5)
```

```{r}
ggplot(penguins, aes(x = body_mass_g, y = 0)) + 
  geom_jitter(width=0, height=0.2, alpha=0.5) + 
  ylim(-1, 5)
```

## Histograms

```{r}
ggplot(penguins, aes(x = body_mass_g)) + 
  geom_histogram()
```

## Density plots

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_density() +
  geom_rug()
```

## Violin plots

```{r}
ggplot(penguins, aes(x = 1, y = body_mass_g)) +
  geom_violin() +
  geom_jitter(height=0, width=0.05, alpha=0.5)
  
```

## Boxplots

```{r}
ggplot(penguins, aes(x = 1, y = body_mass_g)) +
  geom_boxplot() + 
  geom_jitter(height=0, width=0.05, alpha=0.5)
```

## Representing a categorical variable

* Raw data

```{r}
ggplot(penguins, aes(x = body_mass_g, y = 0, color=species)) + 
  geom_jitter(width=0, height=0.2, alpha=0.5) + 
  ylim(-1, 5)
```


* Histogram

```{r}
ggplot(penguins, aes(x = body_mass_g, fill = species)) + 
  geom_histogram(bins=25)
```

Note the above is a stacked histogram, not three different histograms.

To represent each of the species with it's own histogram we need to specifying the position argument as below.

```{r}
ggplot(penguins, aes(x = body_mass_g, fill = species)) + 
  geom_histogram(bins=25, position="identity", alpha=0.75)
```


* Density plots

```{r}
ggplot(penguins, aes(x = body_mass_g, color = species)) + 
  geom_density()
```

* Violin plot w/raw data depicted

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g, color=species)) +
  geom_violin() +
  geom_jitter(height=0, width=0.15, alpha=0.5)
```


* Boxplot w/raw data

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g, color=species)) +
  geom_boxplot(outliers = FALSE) + 
  geom_jitter(height=0, width=0.15, alpha=0.5)
```


# Faceting

Faceting is equivalent to generating plots of the conditional distributions of one or more categorical variable:

```{r}
ggplot(penguins, aes(x = body_mass_g, fill = species)) + 
  geom_histogram(bins=15) + 
  facet_wrap(~species, ncol=1)
```



# Some basic data analysis

`na.rm` require here to remove missing data before calculating means and standard deviations.

```{r}
penguins |>
  summarize(mean_bodymass = mean(body_mass_g, na.rm = TRUE), 
            sd_bodymass = sd(body_mass_g, na.rm = TRUE))
```


Alternate approach is to remove observations with `NA` values

```{r}
penguins |>
  filter(!is.na(body_mass_g)) |>  # filter out the NA values before summarize
  summarize(mean_bodymass = mean(body_mass_g),
            sd_bodymass = sd(body_mass_g))
  
```


If we want the conditional means and standard deviations we can use `group_by` before summarize.  Note the parallel with faceting above.

```{r}
penguins |>
  filter(!is.na(body_mass_g)) |>
  group_by(species) |>
  summarize(mean_bodymass = mean(body_mass_g),
            sd_bodymass = sd(body_mass_g))
```

## ANOVA to test difference in group means

```{r}
anova_bodymass <- aov(body_mass_g ~ species, data = penguins)
anova_bodymass
```

Generate standard ANOVA output table

```{r}
summary(anova_bodymass)
```

# Bivariate plots

-   Scatter plots

Plot the filtered data, using color and shape to represent both species and sex variables simultaneously.

```{r}
ggplot(penguins, aes(x = body_mass_g, 
                     y = flipper_length_mm, 
                     color = species,
                     shape = sex)) + 
  geom_point()
```

# Basic filtering with dplyr

```{r}
penguins_filtered <-
  filter(penguins, !is.na(sex), species != "Gentoo")
```


Notice the legend no longer includes NA, and we've also filtered out all the Gentoo penguins.

```{r}
ggplot(penguins_filtered, aes(x = body_mass_g, 
                     y = flipper_length_mm, 
                     color=species,
                     shape=sex)) + 
  geom_point()
```

# Faceting on multiple variables

```{r}
ggplot(penguins_filtered, aes(x = body_mass_g, 
                     y = flipper_length_mm, 
                     color=species,
                     shape=sex)) + 
  geom_point(alpha=0.5) + 
  facet_wrap(~species) + 
  labs(x = "Body Mass (g)", y = "Flipper length (mm)") +
  theme_classic() +
  theme(aspect.ratio = 1.5)
```

## 2D density plot

```{r}
ggplot(penguins_filtered, aes(x = body_mass_g, 
                     y = flipper_length_mm, 
                     color=species,
                     shape=sex)) + 
  geom_point(alpha=0.75) + 
  geom_density2d(alpha=0.5) +
  facet_wrap(sex~species) + 
  labs(x = "Body Mass (g)", y = "Flipper length (mm)") +
  theme_classic() +
  theme(legend.position = "none",
        aspect.ratio=1)
```

# Grouping on multiple variables to produce summaries of interest

```{r}
penguins_filtered |>
  group_by(species, sex) |>
  summarise(across(c(body_mass_g, flipper_length_mm), 
                   list(mean = mean, sd = sd)))
```
