---
title: "Problem Set 09"
author: ""
date: today
date-format: iso
format: 
  html:
    embed-resources: true
    theme: cosmo
editor: source
---


```{r include = FALSE, silent = TRUE}
library(tidyverse)
```

# Problems
#### 1. Write a for-loop to read each of the files in the poems_corpus directory, returning a list with four character vectors containing: (1) the authors first names; (2) authors last names; (3) poem/song title; (4) text of the poem/song.


```{r}
poem_files <- list.files(path="~/Downloads/poem_corpus", 
                        pattern="*.txt",
                        full.name = TRUE)
poem_files
```

Now we need to write a for loop to extract (1) the authors first names; (2) authors last names; (3) poem/song title; (4) text of the poem/song from the files.

Now we have first three info from the file name. for that we will use basename function to remove path. Then we remove .txt. Then we use strsplit by "_" to splot first name, last name and title. Used collapse in c() to combine things into a single string. if not, each line would become a vector

```{r, warning=FALSE}
first_name <- c()
last_name <- c()
poem_title <- c()
poem_text <- c()

file <- poem_files[1]

for (file in poem_files) {
  file_name_npath <- basename(file)
  file_name_noext <- sub("\\.txt$", "", file_name_npath)

  parts <- strsplit(file_name_noext, "_")[[1]]
  
  first_name <- c(first_name, parts[1])
  last_name <- c(last_name, parts[2])
  poem_title <- c(poem_title, paste(parts[3:length(parts)], collapse = " "))
  
  texts <- paste(readLines(file), collapse = "\n") # used paste because all the lines should be in the same vector
  poem_text <- c(poem_text, texts)
}

poem_list <- list(
  first_name = first_name,
  last_name = last_name,
  poem_title = poem_title,
  poem_text = poem_text
)
```


#### 2. Create an empty data frame with the following column names and types and assign it the variable name poems_df:

first_name (character), last_name (character), title (character), stanza (integer), num_lines (integer), num_words (integer), num_alpha (integer), num_punct (integer)

```{r}
poems_df <- tibble(
  first_name = character(),
  last_name = character(),
  title = character(),
  stanza = integer(),
  num_lines = integer(),
  num_words = integer(),
  num_alpha = integer(),
  num_punct = integer()
)
```

#### 3. Write a for-loop to iterate over the vectors you created in Problem 1 in parallel. Apply the analyze_poem function to the text of each poem, and use mutate to add author and title information to the output of analyze_poem. Merge each resulting data frame into poems_df using the dplyr::bind_rows.

Analyze poem function from previous files

```{r}
get_stanzas <- function(poem) {
  str_split_1(poem, "\\n\\s*\\n")
}

count_lines <- function(stanza) {
  str_count(stanza, "\n") + 1
  
}

count_words <- function(stanza) {
  words <- 
    stanza |> 
    str_squish() |>
    str_split_1("\\s")
  
  length(words)
}

count_alphabetical <- function(stanza) {
  return (str_count(stanza, "[:alpha:]"))
}

count_punctuation <- function(stanza) {
  return (str_count(stanza, "[:punct:]"))
}

####

analyze_poem <- function(poem) { 
  stanzas <- get_stanzas(poem)
  
  nstanzas <- length(stanzas)
  linects <- integer(length = nstanzas)
  wordcts <- integer(length = nstanzas)
  alphacts <- integer(length = nstanzas)
  punctcts <- integer(length = nstanzas)
  
  for (i in 1:nstanzas) {
    stanza <- stanzas[i]
    linects[i] <- count_lines(stanza)
    wordcts[i] <- count_words(stanza)
    alphacts[i] <- count_alphabetical(stanza)
    punctcts[i] <- count_punctuation(stanza)
  }
  
  tibble(lines = linects,
         words = wordcts,
         alpha = alphacts,
         punct = punctcts)
}

```

```{r}
analyze_poem(poem_list[['poem_text']][1])
get_stanzas(poem_list[['poem_text']][1])

```

It seems to be working. Note: I updated line break as "\\n\\n" OR "\\n \\n" in the get_stanzas() as line_breaks were defined that way in my code

For loop - Just found out I dont have to do poem_list['poem_text'], can just use poem_text

```{r}

for (i in 1:length(poem_text)) {
  poem_analysis <- analyze_poem(poem_text[[i]])
  
  poem_analysis <- poem_analysis |> 
    mutate(
      first_name = first_name[i],
      last_name = last_name[i],
      title = poem_title[i],
      stanza = row_number()
    ) |> 
    rename(
      num_lines = lines,
      num_words = words,
      num_alpha = alpha,
      num_punct = punct
    )
  
  poems_df <- bind_rows(poems_df, poem_analysis)
}

poems_df

```

#### 4. Using the output of Problem 3, add the following metrics (per stanza) to the data frame:

word_complexity – the average number of alphanumeric characters per word
punct_density – the average number of punctuation characters per line

```{r}
poems_df |> 
  mutate(
    word_complexity = num_alpha/num_words,
    punct_density = num_punct/num_lines
  )
```

#### 5. On a per poem basis (hint: group_by), generate a derived data frame with the following:
stanza_complexity – the average word_complexity for the poem as a whole
poem_complexity – the number of stanzas times the stanza complexity
punct_metric – the median punct_density for the poem as a whole

```{r}
poems_df |> 
  group_by(title) |> 
  mutate(
    word_complexity = num_alpha/num_words,
    punct_density = num_punct/num_lines
  ) |> 
  summarise(
    stanza_complexity = mean(word_complexity),
    poem_complexity = n()* stanza_complexity,
    punct_metric  = median(punct_density),
  )
```

#### 6. What are the top three poems in terms of poem_complexity? What are the top three poems in terms of stanza_complexity?
Will use arrange and slice for this.

```{r}
poems_df |> 
  group_by(title) |> 
  mutate(
    word_complexity = num_alpha/num_words,
    punct_density = num_punct/num_lines
  ) |> 
  summarise(
    stanza_complexity = mean(word_complexity),
    poem_complexity = n()* stanza_complexity,
    punct_metric  = median(punct_density),
  ) |> 
  arrange(desc(poem_complexity)) |> 
  slice_head(n = 3)
```

In terms of poem_complexity, Anyone-Lived-in-a-Pretty-How-Town, Jabberwocky and Circle-Game	are top 3

```{r}
poems_df |> 
  group_by(title) |> 
  mutate(
    word_complexity = num_alpha/num_words,
    punct_density = num_punct/num_lines
  ) |> 
  summarise(
    stanza_complexity = mean(word_complexity),
    poem_complexity = n()* stanza_complexity,
    punct_metric  = median(punct_density),
  ) |> 
  arrange(desc(stanza_complexity)) |> 
  slice_head(n = 3)
```

In terms of stanza_complexity, Jabberwocky, The-Colossus and The-Tyger are top 3

#### 7. Which author’s poems have the smallest punct_metric?

```{r}
poems_df |> 
  group_by(title) |> 
  mutate(
    word_complexity = num_alpha/num_words,
    punct_density = num_punct/num_lines
  ) |> 
  summarise(
    first_name = first_name,
    last_name = last_name,
    stanza_complexity = mean(word_complexity),
    poem_complexity = n()* stanza_complexity,
    punct_metric  = median(punct_density),
  ) |> 
  arrange(punct_metric) |> 
  slice_head(n = 1) 
```

Maya Angelou’s poem has the smallest punc_metric.

