---
title: "Problem Set XX"
author: "Your Name, NetID"
date: today
date-format: iso
format: 
  html:
    embed-resources: true
editor: source
---


# Instructions

Answer the questions below using code blocks that generate the correct outputs. We encourage you to include explanatory text in your markdown document, however **each of your solutions should show how to generate the answer using R code**. 

Write "robust" solutions wherever possible. A good rule of thumb for judging whether your solution is appropriately "robust" is to ask yourself "If I added additional observations or variables to this data set, or if the order of variables changed, would my code still compute the right solution?"

Make sure your markdown is nicely formatted -- use headers, bullets, numbering, etc so that the structure of the document is clear.

Make sure to "Render" your document before submission to confirm that all code blocks and formatting is as you expect.

When completed, title your markdown file as follows (replace `XX` with the assignment number, e.g. `01`, `02`, etc):

-   `netid-assignment_XX.qmd`

Submit both your markdown file and the generated HTML document via the Assignments submission section on Sakai.



# Problems

### 1. 

Describe problem 1

```{r}
# Your code here
```


### 2. 

Describe problem 2.

Add `error=TRUE` to your code block header if it has an error but you don't want that error to interrupt execution of the rest of your notebook.

```{r, error=TRUE}
# Code that generates an error
```

Provide a written description of how you've tried to diagnose the issues you're running into that generate the error above.

### 3. 

Describe problem 3.

```{r}
# more code here
```

If necessary or useful provide a description of how your code solves the problem.

