---
title: "Bio 724: Strings basics using Stringr"
author: "Paul Magwene"
format: 
  html:
    embed-resources: true
---


## Introduction

In most programming language, strings are the most generic data type, in that most  other data types can be converted to a string representation but only a subset of strings can be converted to the more specialized data types.  

In R, strings are represented by the `character` data type. In this document I will use the term `character` and strings interchangeably. 

As we've done in past lectures, we'll be emphasizing string related functions from the tidyverse, provided by the package ["stringr"](https://stringr.tidyverse.org/). Be aware that there are many base R functions with similar string manipulation functionality; you'll likely encounter such base R functions if you search the web for help.

## Libraries

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```


## Creating strings 

We can create strings on the fly in a variety of different ways:


```{r}
s1 <- 'This is a string'       # single quotes
s2 <- "This is also a string"  # double quotes

# Having both quoting styles allowing for nesting
s3 <- "You can nest 'single quotes' in double quotes"
s4 <- 'She said, "I prefer to nest double quotes in my single quotes"'

s5 <- as.character(3.14)    # string from a double
s6 <- as.character(1 + 1i)  # string from a complex number
s7 <- as.character(TRUE)  # strings from a boolean
```

## String length

Remember that all atomic data types are actually vectors in R:

```{r}
# character vector of length 1
hello <- "Hello, world!"

# character vector of length 
greetings <- c("Hello, world!", "Howdy, pardner!") 
```

So when we apply the `length()` function we get not the number of characters in a string, but rather the number of strings in the vector:

```{r}
length(hello)
length(greetings)
```

To get the number of characters in a string we can use the `stringr::str_length` function:


```{r}
# number of characters in each string in the vector
str_length(hello) 
str_length(greetings)
```



## Special characters

The R character type supports unicode values, so you can create strings with characters from different alphabets or even emojis in them:

```{r}
hello_japan <- "ã“ã‚“ã«ã¡ã¯"
hello_emoji <- "ðŸ‘‹"
```


Some characters, like tabs and newlines lines are "invisible" so we need a special way to represent them. R (and many other programming languages) use "character escaping" to write such characters, as illustrated below:

```{r}
# \t = TAB character
with_tab <- "A line of text\tWith a tab" 

# \n = newline (return)
with_newline <- "A line of text\nWith a newline" 
```

## The `cat` and `print` functions

The cat function, when applied to strings, will print to your screen a representation of the string data with special characters interpretted appropriately:

```{r}
cat(with_tab)

cat(with_newline)
```

The `print` function on the other hand returns a character vector representing the argument(s) passed to it:

```{r}
# contrast the output produce by print to that of cat
print(with_newline)
```

However, print 


An important difference is that `cat` returns a NULL object wile `print` returns the character object.


## Reading a file as a string

The `readr::read_file` function will read all the text in a file as a single string object.


```{r}
seuss <- read_file("cat_in_hat_excerpt.txt")
```

```{r}
seuss
```


```{r}
cat(seuss)
```



## Combining strings

```{r}
str_c(hello, hello_emoji)
```

```{r}
# when you have a vector of strings and 
# you want to combine them into a single string
str_flatten(greetings, collapse = " ")
```

## String interpolation

A very useful function when working with strings is `stringr::str_glue` which provides facilities for "string interpolation".

String interpolation provides a convenient syntax for inserting/interpolation the values of variables in string objects like shown below:


```{r}
# choose a single random integer between 1 and 10
x <- sample(1:100, 1)

choicestr <- str_glue("The integer I chose was {x}!")
choicestr
```



## Simple string manipulation


```{r}
phrase <- "How now, Brown Cow?"

str_to_upper(phrase)

str_to_lower(phrase)

str_to_title(phrase)

str_to_sentence(phrase)
```

Some useful "cleaning operations" include `str_trim` (trim whitespace at ends of string) and `str_squish` (trim whitespace at ends AND within):
```{r}
whitespace <- "   This string has\nlots of      whitespace\taround it and in it.   "

trimmed <- str_trim(whitespace)

squished <- str_squish(whitespace)

print(whitespace)
print(trimmed)
print(squished)
```

```{r}
cat("Original version:\n", whitespace, "\n\n", sep="")
cat("Trimmed version:\n", trimmed, "\n\n", sep="")
cat("Squished version:\n", squished, sep="")
```



## Substrings

```{r}
str_sub(phrase, 5, 7)
str_sub(phrase, 5) # from 5 to end
str_sub(phrase, 5, -2) # from 5 to two from the end
```

`str_sub` can be used with assignment to replace characters:

```{r}
str_sub(phrase, 5, 7) <- "NOW" 
phrase
```

## String splitting

Split a single string. Returns a vector:
```{r}
# takes one string as input as splits on given pattern
str_split_1(phrase, " ") 
```

Apply a splitting operation to multiple strings. Returns a list:
```{r}
phrases <- c("A wolf in sheeps clothing", 
             "That which does not kill us makes us stronger")

# takes a vector of strings as inputs and splits on given pattern
str_split(phrases, " ")
```


Access the splits in the list
```{r}
split_phrases <- str_split(phrases, " ") 
split_phrases[[1]]  # get the splits for the first string in phrases
```


## String matching

```{r}
phrase <- "How now, Brown Cow?"
```

Count the number of times you find a match:
```{r}
str_count(phrase, "ow")
```

Give the indices of the first match:
```{r}
str_locate(phrase, "ow")
```

Give the indices of all the matches
```{r}
str_locate_all(phrase, "ow")
```

Return a string with the first match replaced
```{r}
str_replace(phrase, "ow", "OW")
```

Return a string with all the matches replaced
```{r}
str_replace_all(phrase, "ow", "OW")
```

