---
title: "Strings in R"
format: html
---

## Libraries

```{r}
library(tidyverse)
```


## Basics

In most programming language, strings are typically the most generic data type, in that most  data types can be converted to a string representation but only a subset of strings can be converted to the more specialized data types.  

In R, strings are represented by the `character` data type. In this document I will use the term `character` and strings interchangeably. 

As we've done in past lectures, we'll be emphasizing string related functions from the tidyverse, provided by the package ["stringr"](https://stringr.tidyverse.org/). Be aware that there are many base R functions with similar string manipulation functionality; you'll likely encounter such base R functions if you search the web for help.

## Creating strings 

We can create strings on the fly in a variety of different ways:


```{r}
s1 <- 'This is a string'       # single quotes
s2 <- "This is also a string"  # double quotes

# Having both quoting styles allowing for nesting
s3 <- "You can nest 'single quotes' in double quotes"
s4 <- 'She said, "I prefer to nest double quotes in my single quotes"'

s5 <- as.character(3.14)    # string from a double
s6 <- as.character(1 + 1i)  # string from a complex number
s7 <- as.character(c(TRUE, FALSE))  # strings from a vector of booleans

```


## Special characters

R character objects support unicode values, so you can insert characters from different alphabets or even emojis in them:

```{r}
hello_japan <- "ã“ã‚“ã«ã¡ã¯"
hello_emoji <- "ðŸ‘‹"
```


Some characters, like tabs and newlines lines are "invisible" so we need a special way to represent them. R (and many other programming languages) use "character escaping" to write such characters, as illustrated below:

```{r}
with_tab <- "A line of text\tWith a tab" # the \t = TAB character
with_newline <- "A line of text\nWith a newline" # the \n = newline (return)
```

## The `cat` and `print` functions

The cat function, when applied to strings, will print to your screen a representation of the string data with special characters interpretted appropriately:

```{r}
cat(with_tab)
```

```{r}
cat(with_newline)
```

The `print` function on the other hand returns a character vector representing the arguments passed to it:

```{r}
# contrast the outputs here
cat(with_newline)

print(with_newline)
```

Another important difference is that `cat` returns a NULL object wile `print` returns the character represent.

## Some simple character manipulation


```{r}
phrase <- "How now, Brown Cow?"

str_to_upper(phrase)

str_to_lower(phrase)

str_to_title(phrase)

str_to_sentence(phrase)
```

## Substrings

```{r}
str_sub(phrase, 5, 7)
str_sub(phrase, 5) # from 5 to end
str_sub(phrase, 5, -2) # from 5 to two from the end
```

`str_sub` can be used with assignment to replace characters:

```{r}
str_sub(phrase, 5, 7) <- "NOW" 
phrase
```

## String splitting

```{r}
# takes one string as input as splits on given pattern
str_split_1(phrase, " ") 
```

```{r}
phrases <- c("A wolf in sheeps clothing", 
             "That which does not kill us makes us stronger")

# takes a vector of strings as inputs and splits on given pattern
str_split(phrases, " ")
```

```{r}
split_phrases <- str_split(phrases, " ") 
split_phrases[[1]]
```




## String matching

```{r}
str_count(phrase, "ow")
```

```{r}
str_locate(phrase, "ow")
```

```{r}
str_locate_all(phrase, "ow")
```

```{r}
str_replace(phrase, "ow", "OW")
```

```{r}
str_replace_all(phrase, "ow", "OW")
```




## Reading a file as a string

The `readr::read_file` function will read all the text in a file as a single string object.

```{r}
jabberwocky <- read_file("../../data/jabberwocky.txt")
jabberwocky
```

To see the Jabberwocky text in it's formatted glory, use the `cat` function:

```{r}
cat(jabberwocky)
```

The base R `scan()` function is useful if you want to read in a file as a vector of "words" (where "word" = contigous character delimited by white space):

```{r}
jabberwocky_words <- scan("../../data/jabberwocky.txt", what = character())
jabberwocky_words
```




```{r}
str_length(hello_emoji)
```

```{r}
str_length(hello_japan)
```

```{r}
str_length(jabberwocky)
```



```{r}
str_length(jabbwocky_words)
```


```{r}
# doesn't add up to str_length(jabberwocky); why?
sum(str_length(jabberwocky_words))
```






## String splitting

```{r}
stanzas <- str_split_1(jabberwocky, "\n\n")
stanzas[1]
```



```{r}
stanza2 <- stanzas[2]
lines2 <- str_split_1(stanza2, "\n")
lines2
```

```{r}
trimmed_lines2 <- str_trim(lines2)
trimmed_lines2
```

```{r}
stanza2_clean <- str_flatten(trimmed_lines2, collapse=" ")
stanza2_clean
```

```{r}
stanza2_clean |>
  str_split_1(" ") 
```

```{r}
wordct2 <-
  stanza2_clean |>
  str_split_1(" ") |>
  length()
```

```{r}
stanza2_clean |>
  str_remove_all(" ") |>
  str_length()
```




```{r}
words_per_line <- str_split(trimmed_lines,
                            pattern="[:blank:]")
```






## Working with substrings

```{r}
str_sub(stanza2, 10, 20)
```


## String matching and replacing

```{r}
str_count(jabberwocky, "Jabberwock")
```



```{r}
# str_locate returns the first match
str_locate(jabberwocky, "Jabberwock")
```

```{r}
# str_locate_all returns all matches PER input string
str_locate_all(jabberwocky, "Jabberwock")
```

```{r}
# str_locate_all returns all matches PER input string
str_locate_all(stanzas, "Jabberwock")
```



```{r}
hit <- str_locate(stanza2, "Jabberwock")
hit
```

```{r}
str_sub(stanza2, hit[1], hit[2])
```



```{r}
word <- "Jabberwock"
new_stanza2 <- str_replace(stanza2, word, str_to_upper(word))
cat(new_stanza2)
```

```{r}
jabberwocky |> 
  str_replace_all("Jabberwock", "JABBERWOCK") |>
  cat()
```




## Substrings

```{r}
str_sub()
```



