---
title: "Merges and joins in R"
author: "Greg Wray"
date: today
date-format: iso
format: 
  html:
    embed-resources: true
editor: source
---

## Set up

Load the tidyverse package and data sets:

```{r include = FALSE, silent = TRUE }
library(tidyverse)

# load full weather data set and location metadata
wdata_full <- read_csv("weatherAUS.csv")
wlocs <- read_csv("weather_locations.csv")
```

## Tiny data frames

To illustrate how merges and joins work, we will use two tiny data frames generated by the code below. Note that these correspond to the example data frames in the dplyr cheat-sheet.

```{r tiny-data-frames}
# create tiny data frames 
A <- c("a", "b", "c")                      # generate the columns
A2 <- c("a", "b", "d")
B <- c("t", "u", "v")
B2 <- c("t", "u", "w")
C <- c(1, 2, 3)
D <- c(3, 2, 1)
x <- data.frame(A = A, B = B, C = C)       # generate data frames containing the columnns
y <- data.frame(A = A2, B = B2, D = D)
x                                          # check your work!
y
```


## Weather data

For real data, we'll work with daily weather records for ~2,600 consecutive days from 49 locations in Australia. The last two columns note whether it rained on the day the records were collected and whether it rained on the following day. The goal is to use multi-year weather records to predict rainfall. 

**Data source and modifications.**

Weather data were compiled by Joe Young and Adam Young. Data were downloaded from: https://www.kaggle.com/datasets/jsphyg/weather-dataset-rattle-package.

Below is code to pre-process the full weather data set. This code subsets the data to reduce its size and separates the last column so that we can practice merges. 

```{r pre-process weather data}
# subset data from 6 locations and save
locs <- c("WaggaWagga", "Tuggeranong", "Ballarat", "Mildura", "Cairns", "MountGambier")
wdata6 <- filter(wdata_full, Location %in% locs)
write_csv(wdata6, "wdata_6.csv")
# separate the last column 
wdata6_next <- select(wdata6, RainTomorrow)
wdata6 <- select(wdata6, -RainTomorrow)

# subset data from 1 location and save
wdata1 <- filter(wdata_full, Location == "Uluru")
write_csv(wdata1, "wdata_1.csv")
# separate the last column 
wdata1_next <- select(wdata1, RainTomorrow)
wdata1 <- select(wdata1, -RainTomorrow)

# remove the full data set from memory
rm(wdata_full)
```

Let's start by taking a look at the data. Click on `wdata6` in the Environment pane.

Plots are a good way to explore what the data can tell us.

```{r}
ggplot(wdata6, aes(x=MaxTemp, fill=Location)) +
  geom_histogram(na.rm=TRUE, binwidth=1) +
  facet_wrap(~ Location) +
  theme_light() 
```

```{r}
ggplot(wdata6, aes(x=Humidity9am, y=Humidity3pm, color=Location)) +
  geom_point(na.rm=TRUE, size=0.1, alpha=0.2) +
  facet_wrap(~ Location) +
  theme_light() 
```

```{r}
ggplot(wdata6, aes(x=RainToday, y=WindGustSpeed, color=RainToday)) +
  geom_point(position='jitter', alpha=0.2, size=0.1, na.rm=TRUE) +
  geom_violin(alpha=0.5, na.rm=TRUE) +
  theme_light() 
```

```{r}
ggplot(wdata6, aes(x=RainToday, y=(Pressure9am - Pressure3pm), color=RainToday)) +
  geom_point(position='jitter', alpha=0.2, size=0.1, na.rm=TRUE) +
  geom_violin(alpha=0.5, na.rm=TRUE) +
  facet_wrap(~ Location) +
  theme_light() 
```


## Merge

Merge operations allow you to "paste together" rows or columns from two data frames. Think of this as simply attaching two data frames along one dimension or the other. For this to make sense with your data, the two data frames should have a similar composition in the dimension not being merged: rows should match for column merges, and columns should match for row merges.

Although merges are simpler to understand than joins, they are *much* less powerful and versatile. In addition, merge operations don't consider values within the tables, making them inherently risky and much more likely to produce incorrect results. In practice, only use merges for simple situations where you know for sure that the values in the two data frames will match up in a meaningful way after the merge. Even in these cases, make a habit of checking carefully that the result is what you intended.

**Column merge** 

The simplest way to combine values from two data frames into one is to use `bind_cols()`. This function attaches two data frames together "horizontally". The order of data frame names passed to the function matters: the first data frame will be on the "left" and the second one on the "right"; column order within each will be preserved in the output. 

```{r}
# simple column merge

```
Note that if the two data frames contain one or more columns with the same name, they are are both retained, but they are automatically renamed by appending '...' and the column number. This can get very confusing with larger data frame. Best practice is to provide distinct column names *before* merging. 

```{r}
# rename columns before merging 

```

Now, with the weather data, merge `wdata6_next`, which contains the column `RainTomorrow`, with the main data set:

```{r c-merge}

```                               

Horizontal merges are risky. If rows in the two source data frames are not aligned correctly, you will associate the wrong values in the merged data frame. Below is a contrived example: first randomize values and then merge them. They values are meaningless, but `bind_cols` has no mechanism for detecting this and does not return an error or warning. It's very easy to make this kind of mistake!

```{r c-merge randomized}

```

Another potential problem to be aware of is that the number of rows must be the same in both data frames. If row numbers are mismatched, dplyr will *not* use recycling to make their length equal, and will instead throw an error.  

```{r c-merge unequal}
# attempt to merge data frames with unequal column lengths

```

Checking that a horizontal merge is working correctly is time-consuming and error-prone. A left join (discussed below) is *far* safer than a horizontal merge, and should be the preferred method whenever possible. 

**Row merge**  

Row merges are slightly more complicated than column merges. The `bind_rows()` function attaches two data frames together "vertically". In contrast the horizontal merges, the number of rows does not matter, but the number columns and their data types should match.

Visualize this with tiny data frames:

```{r}
# first, we need to make a new table y to match the slides and cheatsheet
A <- c("c", "d")
B <- c("v", "w")
C <- c(3, 4)
y <- data.frame(A = A, B = B, C = C)

# visualize before merging
x
y
```

Now, carry out the row merge:

```{r}
# simple row merge

```
Note that rows appear in the order provided, and that duplicated rows are retained.

The code below attaches weather observations for an additional location to the data frame for the six locations that we have been using. 

```{r r-merge}
# example of a row merge; rows of wdata1 are appended "below" those of wdata7

```

Show unique locations and count rows to confirm that the operation worked as expected.

```{r r-merge confirm}

```

An optional argument allows you to track which original data frame each rows came from. 

```{r r-merge source}
# you can use any name for this new column

```

It is also possible to perform set operations during vertical merges. For instance, if you want to ensure that there are no duplicated rows in the merged data frame, use `union()` instead of `bind_rows()`.

```{r r-merge union}
# to remove any rows that duplicate values in previous rows

```

To return only rows that are present in both data frames, use `intersect()`. We will create two small data frames with overlapping dates; to make things easier to follow, row numbers in these data frames correspond to date in the month.

```{r r-merge intersect}
# make two small data frames with partially overlapping row numbers

# return only rows present in both (use dates to confirm)

```

It is also possible to return only set differences using `set_difference()`; see documentation for details.

A final cautionary note before moving on: row merges do *not* require an equal number of columns. This is an important difference from column merges. Missing values are replaced with `NA`s. Beware: row merges are very strict with column names, and even minor spelling differences will be treated as different columns, with unwanted consequences! 


## Join

Like merges, joins allow you to combine data from two data frames. However, joins work quite differently from merges. During a join operation, values in a specified *join column* in each data frame are compared; whether these values match determines what ends up in the resulting joined data frame. An explicit set logic is used to specific how the values in the join column(s) should be interpreted. (It is also possible to use multiple join columns, where values must match in all columns before applying set logic.)    

While joins are initially more challenging to understand than merges, they are remarkably powerful and useful. In addition, joins are safer than merges (when used correctly), because they are based on the  values in the data frame rather than position or column name. Joins are used commonly and in diverse situations, so it's important to become familiar with how they work and what they can do for you. dplyr offers seven different join functions, but most are rarely used. We will work with the three most common operations here: left, inner, and full joins. As you become comfortable applying these joins, you will learn to identify situations where one of the other joins might be useful.

**Left joins**

A left join compares values the join columns of each data frame. If the values match, the entire row of the second ("right") data frame is appended onto that specific row of the first ("left") data frame. See the left join slide for a visual and algorithmic description of how a left join works.

We can visualize this with tiny data frames. First, regenerate the tiny data frames x and y using the code below (this is a copy of an earlier code chunk, repeated here so you don't have to scroll up to the top several times).

```{r}
# create tiny data frames
A <- c("a", "b", "c")                      # generate the columns
A2 <- c("a", "b", "d")
B <- c("t", "u", "v")
B2 <- c("t", "u", "w")
C <- c(1, 2, 3)
D <- c(3, 2, 1)
x <- data.frame(A = A, B = B, C = C)       # generate data frames containing the columnns
y <- data.frame(A = A2, B = B2, D = D)
x
y
```

Now, try the join.

```{r}
# default join syntax is simple

# for a left join, order matters

```

Left joins always return at least the number of rows in the first data frame, and may return more. The order of the data frames matters when specifying a left join, since values are compared asymmetrically, from the second to the first (from right to left). This is not the case for some other join methods, where values are compared in both directions. Importantly, matching rows do not need to be in the same order and do not need to be present 1:1 in the two tables. 

A common use case for left joins is when you have a data frame and want to add more columns to each row based on information in a different data frame. For the weather data, we will use a left join to add metadata about each location. Note that each location is present in thousands of rows, and that the exact number of rows for each location is distinct. Adding the location metadata by hand would be clumsy, laborious, and error prone. A join appends the correct information reliably, quickly, and reproducibly. 

```{r}
# append location metadata to the weather data frame

```

To view a random sample of rows, use `slice_sample()` and pass the number of rows you want back.

```{r}
# retrieve a sample of 10 rows

```

By default, all of the dplyr join functions will use all columns with the same name in the two data frames as join columns. This is usually, but not always, what you want. Best practice is to specify the join column(s) explicitly for any code that you want to keep, since it signals what you intend rather than what dplyr assumes by default. 

```{r}
# explicitly specify the join column 

```

**Inner joins**

The purpose of an inner join is to return the intersection set from the join column or columns. In other words, an inner join returns only those rows where there is an identical value in the join column or columns in both tables; rows without a match in the other data frame are discarded. See the inner join slide for a visual and algorithmic description of how an inner join works.

Again, we can visualize this with tiny data frames. 

```{r}
# inner join syntax

# order does not matter for an inner join 

```

During an inner join, no new missing values are generated. This is in contrast with left and full joins. (If there are any missing values *before* an inner join, they are passed on.) 

Now try an inner join with the weather data sets.

```{r}
# inner join syntax

```

Inner joins produce the smallest (or equal to the smallest) return of the three join methods we will cover in class. This is because values in the join column must match in both directions. This also means that order in which data frames are specified in the join statement does not change what is returned. 

**Full joins**

A full join returns every row of both data frames, only joining rows that have a match in the join column(s). Rows without a match are included, and columns present only the other data frame are filled with `NA` values. For this reason, full joins often generate a lot of missing values. Because values in the join column are compared in both directions, order does not matter. See the full join slide for a visual and algorithmic description of how an full join works.

We can again visualize this with the tiny data frames.

```{r}
# full join syntax

# order does not matter for a full join

```

Now try this with the weather data:

```{r}
# full join syntax

```

Because this is a big data set, it's difficult to visualize what just happened. Filtering for rows where the Location matches or does not match can help.

```{r}
# retrieve a row where the join values match

# retrieve a row where the join values do not match

```


**Practicing joins**

We will use the Spellman data set to illustrate a more specific data analysis problem. Our objective is to annotate each gene with some information about what it does. This is a typical use case for a left join.

We can extract information about each gene from the [Gene Ontology database](https://geneontology.org) (GO). This resource uses a controlled vocabulary to describe the product of a gene in three categories: cellular component, molecular function, and biological process. The file `go_slim_mapping.tab` contains GO terms for most genes in the *S. cerevisiae* genome. Both this GO file and the Spellman data use the same set of names for genes. This allows us to use gene ID as the join column to link information between the two data frames.  

```{r}
# import the two data sets
spellman <- read_tsv("spellman-1998-expression.tsv")
GO <- read_tsv("go_slim_mapping.tab", col_names = FALSE)

# prepare the GO data frame: remove unnecessary columns, provide column names, and pivot wider
GO <- select(GO, X1, X2, X4, X5, X7)
GO <- rename(GO, gene_ID = X1, gene_name = X2,GO_cat = X4, GO_term = X5, gene_type = X7)
GO_wide <- pivot_wider(GO, names_from = GO_cat, values_from = GO_term)
# note the somewhat awkward list when multiple GO terms exist for a single category

# now add some informative names to the GO columns
GO_wide <- rename(GO_wide, cellular_component = C, molecular_function = F, biological_process = P)

# perepare the Spellman data frame: repair the first column name
spellman <- rename(spellman, "gene_ID" = "...1")   # using quotes because the current name is weird

# left join to add GO terms to every gene
spellman_GO_annotated <- left_join(spellman, GO_wide, by = join_by(gene_ID))
glimpse(spellman_GO_annotated)
```

Often, it is helpful to reorganize and trim the result of a large join like the one we just did. The code below first rearranges the order of the gene name and GO columns to be near the gene_ID column, then removes columns unrelated to the "alpha" experiment.

```{r}
# move annotations to the left side of the data frame
spellman_GO_annotated <- spellman_GO_annotated |> 
  relocate(gene_name, gene_type, .after = gene_ID) |> 
  relocate(cellular_component, molecular_function, biological_process, .after = gene_type)

# remove columns not related to the "alpha" experiment 
spellman_GO_trim <- spellman_GO_annotated |> 
  select(!(starts_with("cl"))) |> 
  select(!(starts_with("cdc"))) |> 
  select(!(starts_with("elu"))) 

# now we can focus on just one experiment with annotations regarding each gene's function
glimpse(spellman_GO_trim)
```

